Detection and replication of epistasis influencing transcription in humans

### 2013-09-02 docs/manuscript/methods.tex
minor edits to the re-written manuscript and methods

\begin{equation}
y = \mu + \beta_{1}c + \beta_{2}p + \beta_{3}s + \beta_{4}a + g + e
\label{eq:lm}
\end{equation}
where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively, fitted as fixed effects; and $g$ is a random additive polygenic effect with a variance covariance matrix 
\begin{equation}
G_{jk} = \left \{ 
\begin{array}{ll}
\sigma _a ^2             &          j = k \\ 
2\phi _{ijk} \sigma _a ^2& 			j \neq k \\
\end{array} \right.
\end{equation}
The parameter $\sigma_a^2$ is the variance component for additive background genetic. Here, we are using family based pedigree information rather than SNP based IBD to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$. The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero. Residuals were normalised by rank transformation and used as the adjusted phenotype for the pairwise epistasis scan to remove any skewness and avoid results being driven by outliers. The GenABEL package for R was used to perform the normalisation \cite{Aulchenko2007}.

@@

@@

Following filtering on this threshold only SNP pairs with at least 5 data points in all 9 genotype classes were kept. We then calculated the LD between interacting SNPs (amongst unrelated individuals within the discovery sample and also from 1000 genomes data) and removed any pairs with $r^2 > 0.1$ or $D'^{2} > 0.1$ to avoid the inclusion of haplotype effects and to increase the accuracy of genetic variance decomposition. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained. At this stage 6404 filtered SNP pairs remained (filter 1). We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed, creating a second set of 4751 filtered SNP pairs (filter 2).

@@

All variance decompositions were calculated using the NOIA method \cite{Alvarez-Castro2008}.

@@

@@

The probability of observing the number of concordant signs is calculated using a binomial test.

>>>>

\begin{equation}
y = \mu + \beta_{1}c + \beta_{2}p + \beta_{3}s + \beta_{4}a + g + e
\label{eq:lm}
\end{equation}
where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively, fitted as fixed effects; and $g$ is a random additive polygenic effect with a variance covariance matrix 
\begin{equation}
G_{jk} = \left \{ 
\begin{array}{ll}
\sigma _a ^2             &          j = k \\ 
2\phi _{jk} \sigma _a ^2& 			j \neq k \\
\end{array} \right.
\end{equation}
The parameter $\sigma_a^2$ is the variance component for additive background genetic. Here, we are using family based pedigree information rather than SNP based IBD to account for relationships between individuals and so $\phi _{jk}$ is the kinship coefficient between individuals $j$ and $k$. The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero. Residuals were normalised by rank transformation and used as the adjusted phenotype for the pairwise epistasis scan to remove any skewness and avoid results being driven by outliers. The GenABEL package for R was used to perform the normalisation \cite{Aulchenko2007}.

@@

\subsubsection{Stage 1}

@@

\subsubsection{Filtering}

We used two approaches to filter SNPs from stage 1 to be tested for significant interaction effects in stage 2.

\paragraph{Filter 1} After keeping SNP pairs that surpassed the $2.91 \times 10^{-16}$ threshold in stage 1 only SNP pairs with at least 5 data points in all 9 genotype classes were kept. We then calculated the LD between interacting SNPs (amongst unrelated individuals within the discovery sample and also from 1000 genomes data) and removed any pairs with $r^2 > 0.1$ or $D'^{2} > 0.1$ to avoid the inclusion of haplotype effects and to increase the accuracy of genetic variance decomposition. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained. At this stage 6404 filtered SNP pairs remained.

\paragraph{Filter 2} We also performed a second filtering screen applied to the list of SNP pairs from stage 1 that was identical to filter 1 but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed \cite{Powell2013}, creating a second set of 4751 unique filtered SNP pairs.

\subsubsection{Stage 2}

@@

@@

\subsection{Population stratification}

\subsection{Probe mapping}

To avoid possibility that epistatic signals might arise due to expression probes hybridising in multiple locations we verified that probe sequences for genes with significant interactions mapped to only a single location. As an initial verification we performed a BLAST search of the full probe sequence against 1000 genomes phase 1 version 3 human genome reference and ensured that only one genomic location aligned significantly ($p < 0.05$). As a second step, to mitigate the possibility of weak hybridisation elsewhere in the genome we divided the probe sequence into three sections (1-25bp, 13-37bp, 26-50bp) and performed a BLAST search of these probe sequence fragments. No probe sequemces or probe sequence fragments mapped to positions other than the single expected genomic target ($p < 0.05$).

@@

The probability of observing the number of concordant signs is calculated using a binomial test. All variance decompositions were calculated using the NOIA method \cite{Alvarez-Castro2008}.

### 2013-09-02 docs/manuscript/nature_letter.tex
minor edits to the re-written manuscript and methods

Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome. The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and examples have been reported in other species \cite{Carlborg2006}, few convincing examples with independent replication exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$). Replication of these signals in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31})$ and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 434$. There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of natural genetic polymorphisms interacting to influence human traits.

@@

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions. First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood (5\% significance threshold $ p < 2.91 \times 10^{-16}$, Methods). Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $\alpha \approx 0.07$, Methods, Supplementary Figure \ref{fig:type1}). Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}). Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/434$, Table \ref{tab:bonferroni}). These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}). We decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A \times A$, $A \times D$, $D \times A$, $D \times D$) and tested for concordance of the sign of the most signicant effect in the independent datasets (Methods). The most significant epistatic effect in the discovery sample was in the same direction in both replication datasets in 22 out of the 30 significantly replicated interactions ($p = 3.76 \times 10^{-8}$, Supplementary Table \ref{tab:sign}).

>>>>

Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome. The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and examples have been reported in other species \cite{Carlborg2006}, few convincing examples with independent replication exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$). Replication of these signals in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 434$. There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of natural genetic polymorphisms interacting to influence human traits.

@@

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions. First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood (5\% significance threshold $ p < 2.91 \times 10^{-16}$, Methods). Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $\alpha \approx 0.07$, Methods, Supplementary Figure \ref{fig:type1}). Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}). Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/434$, Table \ref{tab:bonferroni}). To quantify the similarity of GP maps between the independent datasets (Figure \ref{fig:gpmaps}) we decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A \times A$, $A \times D$, $D \times A$, $D \times D$) and tested for concordance of the sign of the most signicant effect (Supplementary Table \ref{tab:sign}, Methods). Sign concordance between the discovery and both replication datasets was observed in 22 out of the 30 significantly replicated interactions (expected value $= 7.5$ under the null hypothesis of no interactions, $p = 3.76 \times 10^{-8}$).
